Bootstrapping Multi-atlas Segmentation Using Multiple Automatically Generated Templates for the Segmentation of the Whole Hippocampus and Subfields

### 2014-02-08 paper.Rnw
safety checkin'

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-xval-load,include=FALSE,cache=TRUE>>=
jens_atlases = read.csv('data/jens_atlases.csv')
jens_xval_data  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data.csv.gz'))
jens_xval_mean  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data_mean.csv.gz'))
jens_xval_mean$reg_method = factor(jens_xval_mean$reg_method, c("ANTS", "ANIMAL"))
jens_xval_mean$method.mb = factor(jens_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))
mb.kappa.best = round(mean(subset(jens_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 9)$k.mb), digits=3)
mb.diff.best = mean(subset(jens_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 1)$k_diff)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-COMPLETE
% Prepares the data in the form needed for plotting.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
<<ADNI-volumes-prep,echo=F,cache=T>>=
means.complete   <- read.csv('data/cache/ADNI1:qc.csv')
package_totals   <- read.csv('data/cache/ADNI1:package_totals.csv')
@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: prep
% Prepares the data in a the form needed for plotting.
<<ADNI1-SNT-xval-prep,include=FALSE,cache=TRUE>>=
snt_xval_data  <- read.csv(gzfile('data/cache/ADNI-SNT-XVAL:all_data.csv.gz'))
snt_xval_mean  <- read.csv(gzfile('data/cache/ADNI-SNT-XVAL:all_data_mean.csv.gz'))
snt_xval_mean$reg_method = factor(snt_xval_mean$reg_method, c("ANTS", "ANIMAL"))
snt_xval_mean$method.mb = factor(snt_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))

snt.mb.kappa.best =round(mean(subset(snt_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 9)$k.mb), digits=3)
snt.mb.diff.best = mean(subset(snt_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 1)$k_diff)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{color, url, amsmath, geometry, ctable}
\usepackage{titlesec, siunitx, graphicx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}   % keep floats in their place.
\usepackage{authblk}             % Listing Author affiliations
\usepackage{rotating}            % rotating figures
\usepackage[hypcap]{caption}
\usepackage{subcaption}
\newcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2}}
\usepackage{lineno}              % line numbers
\usepackage{multirow,rotating}   % vertical text in tables
\usepackage[displaymath, tightpage]{preview} %draft mode
\setlength{\marginparwidth}{1.2in}
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,RE]{draft - January, 2014}

% styling
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}
\renewcommand\Affilfont{\itshape\small}  %authblk
\linespread{1.3} % one and a half line spacing

@@

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Min Tae M. Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4,5]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,7,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.ucla.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.ucla.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

We conduct two separate Monte Carlo cross-validation experiments comparing \mb{}
and \multiatlas{} whole hippocampal segmentation using differing atlas and
template library sizes, and registation and label fusion methods. The first
experiment is a 10-fold validation (per parameter setting) over 60 subjects
taken from the ADNI Alzheimer's disease imaging database, and the second is a
5-fold validation over 81 subjects having had a first episode of psychosis. In
both cases, automated segmentations are compared with manual segmentions
following the \prsnr{}-protocol. Using the best settings found from these
experiments, we segment \Sexpr{length(means.complete$RID)} images of the
\adnidataset{} dataset and compare these with segmentations from existing
automated methods: \fsl{}, \freesurfer{}, \maper{}, and \snt.  Finally, we
conduct a leave-one-out cross-validation (LOOCV) of hippocampal subfield
segmentation using five high-resolution manually segmented atlases
\citep{Winterburn2013}.

@@

  \item \expfepxval{} consists of a similar cross-validation to explore \mb{}
  segmentations on the brain images of young, first episode psychosis patients,
  and to establish parameter settings that do not overfit on a single dataset.

  \item \expadni{} bridges \mb{} with the existing segmentation literature by
  comparing \mb{} whole hippocampus segmentations with those of well-known
  automated methods (\freesurfer, \fsl, \maper, \snt) on the entire
  \adnidataset{} image dataset consisting of \Sexpr{length(means.complete$RID)}
  brain images of subjects diagnosed as cognitively normal, having mild cognitive
  impairment, or Alzheimer's disease.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.ucla.edu) between March 2012 and
August 2012. The image dataset used was the \adnidataset{} standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.usc.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the ``best'' after quality control. All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citet{Jack2008}. Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: show variability of MAGeT over parameters
<<fig:ADNI1-xval-variability,dependson='ADNI1-Xval-prep',fig.show='hide'>>=
ants = subset(jens_xval_data, 
                     reg_method=="ANTS" & 
                     method.mb=="Majority Vote" &
                    templates.mb %in% seq(1,20,2))

@@

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} and two other popular segmentation packages, \fsl{} and
\freesurfer{}. Firstly, \mb{} is used with the five Winterburn atlas images and
labels as atlases, and a randomly selected subset of 19 psychosis dataset images
as templates. Secondly, \mb{} is run using the same template images, but instead
as atlases we use five additional first episode psychosis subjects and
corresponding manual segmentations (not included above). Finally, \fsl{} and
\freesurfer{} are run with their default settings to segment the entire dataset.
The \fsl{} \verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer{} was run using \verb+recon-all -all+.

@@

As in \expadnixval, we find that similarity score increases with a greater
number of atlases or templates but quickly plateaus (Figure \ref{fig:FEP-xval}).
A maximum similarity score of \Sexpr{snt.mb.kappa.best} is found when using 9
atlases, 19 templates, ANTS registration, and majority vote label fusion.

@@

As \citet{Bland1986} noted, high correlation amongst measures of the same
quantity does not necessarily imply agreement (as correlation can be driven by a
large range in true values, for instance). Figure \ref{fig:FEP-Bland-Altman}
shows Bland-Altman plots illustrating the level of agreement of each method with
manual volumes. All methods show an obvious proportional bias: \freesurfer{}
and \fsl{} markedly underestimate smaller hippocampi and over-estimate large
hippocampi (the limits of agreement are between $-2482^3mm$ and $-784^3mm$, and
between $-1653^3mm$ and $79^3mm$, respectively), whereas both \mb{} methods show
a much less exaggerated, but conservative bias (limits of agreement between
$-67^3mm$ and $766^3mm$ when using FEP atlases, and between $-333^3mm$ and
$504^3mm$ when using Winterburn atlases). On average, \freesurfer{} and \fsl{}
overestimate hippocampal volume by about $1600 mm^3$ and $800 mm^3$,
respectively. In contrast, on average \mb{} underestimates volumes by about $300
mm^3$ when using FEP atlases, and by about $80mm^3$ when using Winterburn
atlases.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FEP-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
<<fig:FEP-xval,fig.show='hide'>>=
  fep.xval <-read.csv('data/results-fep-xval-ants-2013-12-29-191914.csv')
  fep.xval$atlases <- as.factor(fep.xval$atlases)
  fep.xval.mean <- ddply(fep.xval, .(subject,atlases,templates,label), function (df) {
    data.frame(k=mean(df$k),volume=mean(df$volume))
  })

  fep.mb.kappa.best = round(mean(subset(fep.xval.mean, 
                                 templates == 19 & 
                                 atlases      == 9)$k), digits=3)
  stderr <- function(x) sqrt(var(x)/length(x))
  ggplot(fep.xval.mean, aes(x=templates,y=k, colour=atlases)) + 
    stat_summary(fun.y=mean,geom='line', aes(y=k,weight=1,group=atlases)) + 
    stat_summary(fun.y=mean,geom='point',aes(y=k,weight=1,group=atlases)) + 
    stat_summary(fun.y=mean,
                 fun.ymin=function (x) mean(x) - stderr(x),
                 fun.ymax=function (x) mean(x) + stderr(x), 
                 geom='errorbar',aes(y=k, weight=1, group=atlases), 
                 linetype='solid', width=0.5) + 
    #stat_smooth(method='lm', formula=y~log(x)) + 
    scale_y_continuous(breaks=seq(0,1,0.01)) + 
    scale_colour_hue(name="Number of Atlases") +
    scale_linetype(name="Registration Method") +
    xlab( "Number of Templates" ) + 
    ylab( "Mean similarity (DSC)" ) + 
    theme(legend.direction = "horizontal", legend.position = "bottom") 
@

@@

<<fig:FEP-Bland-Altman,dependson='fig:FEP-vols',fig.width=10,fig.height=3,fig.show='hide'>>=
fep.volumes$diff = fep.volumes$manual - fep.volumes$volume
fep.volumes$mean = (fep.volumes$manual + fep.volumes$volume)/2
limits = ddply(fep.volumes, c("Method"), function (df) { 
  data.frame(
    y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T) 
)})

ggplot(fep.volumes, aes(x= mean, y = diff)) + 
  facet_wrap( ~ Method, nrow=1, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  geom_point(size=1) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean manual and computed volume ' (mm^3))) + 
  ylab(expression('manual - computed volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.48\textwidth} 
    \includegraphics[width=\textwidth]{figure/fig:FEP-xval}
    \caption{Dice's similarity score vs. atlas and template library size} 
    \label{fig:FEP-xval}
  \end{subfigure}%
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:FEP-vols} 
    \caption{Computed vs. manual hippocampus volume} 
    \label{fig:FEP-vols}
  \end{subfigure}
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:FEP-Bland-Altman}
    \caption{Bland-Altman plots of computed vs. manual hippocampus volume} 
    \label{fig:FEP-Bland-Altman}
  \end{subfigure}
  \caption{\textbf{First Episode Patient dataset validation.}
Manual segmentation of all 81 subjects is completed using the \prsnr-protocol.
\mb{} is configured using \ants{} registration algorithm and majority vote label
fusion. 
(\ref{fig:FEP-xval}) shows mean DSC overlap similarity score of \mb{}
  segmentations as they vary with atlas and template library size over 5-fold
  validation. Error bars indicate standard error. 
(\ref{fig:FEP-vols}) shows the volumes of the 81 subjects as determined by \fsl,
  \freesurfer, and \mb{} configured in two ways: with an atlas library
  of the five Winterburn atlases (MAGeT-Winterburn), or with an atlas
  library of five manually segmented FEP subjects as atlases (MAGeT-FEP). Linear
  fit lines are shown, with the shaded region showing standard error.
(\ref{fig:FEP-Bland-Altman}) shows the agreement between computed volumes and
  manually segmented volume. The overall mean difference in volume, and limits of
  agreement ($\pm 1.96SD$) are shown by dashed horizontal lines. Linear fit lines
  are shown for each diagnosis group. Note, points below the mean difference
  indicate overestimation of the volume with respect to the manual volume, and
  vice versa.}
\end{figure}

@@

\paragraph{\adnidataset{} dataset}  %%%%%%%%%%%
The {\em \adnidataset{}} standardized dataset contains $1919$ images in total.
\snt{}, \maper{}, and \freesurfer{} hippocampal volumes for a subset of images were
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

See Section \ref{adnidataset} for study details, inclusion criteria and imaging
characteristics.

@@

\paragraph{Winterburn Atlases}  %%%%%%%%%%
The Winterburn atlases \citep{Winterburn2013} are digital hippocampal
segmentations of five in-vivo $0.3mm$-isotropic T1-weighted MR images, described
more fully in \expfepxval, section \ref{exp:fep-xval}.

\paragraph{Experiment details} %%%%%%%%%%
\mb{} was configured with an atlas library composed of the five Winterburn atlas
images and segmentations. A template library of 19 images were randomly selected
from the target dataset of ADNI subjects, and \ants{} registration and majority
vote label fusion were used as these were found to perform favourably in earlier
experiments.

@@

%%%%%%%%%
% ADNI volumes by automated method
%%%%%%%%%
<<fig:ADNI-volumes-plot,dependson='ADNI-volumes-prep',fig.show='hide'>>=
melted=melt(means.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "Method", value.name = "volume")

>>>>

<<load,include=FALSE,cache=TRUE>>=

########################################
# ADNI1-XVAL: prep
# Prepares the data in a the form needed for plotting.
jens_atlases = read.csv('data/jens_atlases.csv')
jens_xval_data  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data.csv.gz'))
jens_xval_mean  <- read.csv(gzfile('data/cache/ADNI-JENS-XVAL:all_data_mean.csv.gz'))
jens_xval_mean$reg_method = factor(jens_xval_mean$reg_method, c("ANTS", "ANIMAL"))
jens_xval_mean$method.mb = factor(jens_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))
mb.kappa.best = round(mean(subset(jens_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 9)$k.mb), digits=3)
mb.diff.best = mean(subset(jens_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 1)$k_diff)

########################################
# ADNI1-COMPLETE
# Prepares the data in the form needed for plotting.
########################################
means.complete   <- read.csv('data/cache/ADNI1:qc.csv')
package_totals   <- read.csv('data/cache/ADNI1:package_totals.csv')

########################################
# ADNI1-SNT-XVAL: prep
# Prepares the data in a the form needed for plotting.
snt_xval_data  <- read.csv(gzfile('data/cache/ADNI-SNT-XVAL:all_data.csv.gz'))
snt_xval_mean  <- read.csv(gzfile('data/cache/ADNI-SNT-XVAL:all_data_mean.csv.gz'))
snt_xval_mean$reg_method = factor(snt_xval_mean$reg_method, c("ANTS", "ANIMAL"))
snt_xval_mean$method.mb = factor(snt_xval_mean$method.mb, c("Majority Vote", "Cross-correlation Vote", "NMI Vote"))

snt.mb.kappa.best =round(mean(subset(snt_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 9)$k.mb), digits=3)
snt.mb.diff.best = mean(subset(snt_xval_mean, 
                                 reg_method   == "ANTS" & 
                                 method.mb    == "Majority Vote" &
                                 templates.mb == 19 & 
                                 atlases      == 1)$k_diff)

########################################
# ADNI1-FEP-XVAL: prep
# Prepares the data in a the form needed for plotting.
fep.xval <-read.csv('data/results-fep-xval-ants-2013-12-29-191914.csv')
fep.xval$atlases <- as.factor(fep.xval$atlases)
fep.xval.mean <- ddply(fep.xval, .(subject,atlases,templates,label), function (df) {
  data.frame(k=mean(df$k),volume=mean(df$volume))
})

fep.mb.kappa.best = round(mean(subset(fep.xval.mean, 
                               templates == 19 & 
                               atlases   == 9)$k), digits=3)
@
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\documentclass{article}
\usepackage[margin=1in]{geometry}
\usepackage{color, url, amsmath, geometry, ctable}
\usepackage{titlesec, siunitx, graphicx}
\usepackage[round,authoryear]{natbib}
\usepackage[section]{placeins}   % keep floats in their place.
\usepackage{authblk}             % Listing Author affiliations
\usepackage{rotating}            % rotating figures
\usepackage[hypcap]{caption}
\usepackage{setspace}            % line spacing control
\usepackage{subcaption}          % subfigures
%\newcommand{\subfloat}[2][need a sub-caption]{\subcaptionbox{#1}{#2}}
\usepackage{lineno}              % line numbers
\usepackage{multirow,rotating}   % vertical text in tables
%\usepackage[displaymath, tightpage]{preview} %draft mode
\usepackage{fancyhdr}
\pagestyle{fancy}
\fancyhead[RO,RE]{}

% styling
\newcommand{\subsubsubsection}[1]{\paragraph{#1}}
\renewcommand\Affilfont{\itshape\small}  %authblk
\onehalfspacing

@@

% title and authors
%\title{Bootstrapping Multi-atlas Hippocampal Subfield Segmentation by using
%Multiple Automatically Generated Templates}
\title{Bootstrapping Multi-atlas Segmentation Using Multiple Automatically
Generated Templates for the Segmentation of the Whole Hippocampus and
Subfields}
\author[1]{Jon Pipitone}
\author[1]{Min Tae M. Park}
\author[1]{Julie Winterburn}
\author[1,9]{Tristram A. Lett}
\author[2,3]{Jason P. Lerch}
\author[4]{Jens C. Pruessner}
\author[4,5]{Martin Lepage}
\author[1,6,9]{Aristotle N. Voineskos}
\author[1,6,7,8]{M. Mallar Chakravarty}
\author[ ]{the Alzheimer's Disease Neuroimaging Initiative\footnote{Data
used in preparation of this article were obtained from the Alzheimer's Disease
Neuroimaging Initiative (ADNI) database (adni.loni.usc.edu). As such, the
investigators within the ADNI contributed to the design and implementation of
ADNI and/or provided data but did not participate in analysis or writing of this
report. A complete listing of ADNI investigators can be found at:
http://adni.loni.usc.edu/wp-content/uploads/how\_to\_apply/ADNI\_Acknowledgement\_List.pdf}}
\affil[1]{Kimel Family Translational Imaging-Genetics Lab, Centre for Addiction and
Mental Health, Toronto, ON, Canada}
\affil[2]{Neurosciences and Mental Health Laboratory, Hospital for Sick
Children, Toronto, ON, Canada}
\affil[3]{Department of Medical Biophysics, University of Toronto, Toronto, ON,
Canada}
\affil[4]{Douglas Mental Health University Institute, Verdun, QC, Canada}
\affil[5]{Department of Psychiatry, McGill University, Montreal, QC, Canada}
\affil[6]{Department of Psychiatry, University of Toronto, Toronto, ON, Canada}
\affil[7]{Institute of Biomaterials and Biomedical Engineering, University of
Toronto, Toronto, ON, Canada}
\affil[8]{Rotman Research Institute, Baycrest, Toronto, ON, Canada}
\affil[9]{Institute of Medical Science, University of Toronto, Toronto, ON, Canada}

@@

We conduct two separate Monte Carlo cross-validation experiments comparing \mb{}
and \multiatlas{} whole hippocampal segmentation using differing atlas and
template library sizes, and registation and label fusion methods. The first
experiment is a 10-fold validation (per parameter setting) over 60 subjects
taken from the Alzheimer's Disease Neuroimaging Database (ADNI), and the second
is a five-fold validation over 81 subjects having had a first episode of psychosis.
In both cases, automated segmentations are compared with manual segmentions
following the \prsnr{}-protocol. Using the best settings found from these
experiments, we segment \Sexpr{length(means.complete$RID)} images of the
\adnidataset{} dataset and compare these with segmentations from existing
automated methods: \fsl{}, \freesurfer{}, \maper{}, and \snt.  Finally, we
conduct a leave-one-out cross-validation (LOOCV) of hippocampal subfield
segmentation using five high-resolution manually segmented atlases
\citep{Winterburn2013}.

@@

  \item \expfepxval{} is a similar cross-validation to explore \mb{}
  segmentations on the brain images of young, first episode psychosis patients. 
  In addition, \mb{} segmentations with two different atlas segmentation
  protocols are compared to automated segmentations by the \fsl and
  \freesurfer algorithms.  The results of this experiment combined with the
  previous experiment establishes parameter settings that do not overfit to the
  neuroanatomical features of a specific patient cohort.

  \item \expadni{} bridges \mb{} with the existing segmentation literature by
  comparing \mb{} whole hippocampus segmentations with those of several well-known
  automated methods (\freesurfer, \fsl, \maper, \snt) on the entire
  \adnidataset{} image dataset consisting of \Sexpr{length(means.complete$RID)}
  brain images of subjects diagnosed as cognitively normal, having mild cognitive
  impairment, or Alzheimer's disease.

@@

Clinical, demographic and pre-processed T1-weighted MRI were downloaded by the
authors from the ADNI database (adni.loni.usc.edu) between March 2012 and
August 2012. The image dataset used was the \adnidataset{} standardized dataset
available from ADNI \footnote{
\url{http://adni.loni.usc.edu/methods/mri-analysis/adni-standardized-data/}}
\citep{Wyman2012}. This image collection contains uniformly pre-processed images
which have been designated to be the ``best'' after quality control. All images
were acquired using 1.5T scanners (General Electric Healthcare, Philips Medical
Systems or Siemens Medical Solutions) at multiple sites using the protocol
described in \citet{Jack2008}. Representative 1.5T imaging parameters were TR =
2400ms, TI = 1000ms, TE = 3.5ms, flip angle = $8^{\circ}$, field of view = 240 x
240mm, a $192 \times 192 \times 166$ matrix ($x$, $y$, and $z$ directions)
yielding voxel dimensions of $1.25mm \times 1.25mm \times 1.2mm$.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% ADNI1-XVAL: show variability of MAGeT over parameters
<<fig:ADNI1-xval-variability,dependson='load',fig.show='hide'>>=
ants = subset(jens_xval_data, 
                     reg_method=="ANTS" & 
                     method.mb=="Majority Vote" &
                    templates.mb %in% seq(1,20,2))

@@

In addition to the MCCV, we segment the entire first episode psychosis dataset
using \mb{} using two different atlases, as well as two popular automated
segmentation packages, \fsl{} and \freesurfer{}. Specifically, \mb{} is run once
with the five Winterburn atlas images and labels as atlases and a randomly selected
subset of 19 target images as templates. \mb{} is run a second time using
the same template images, but we using five additional first
episode psychosis subjects and corresponding manual segmentations (not included
above) as atlases. \fsl{} and \freesurfer{} are run with the default settings:
\fsl{} \verb+run_first_all+ script was used according to the FIRST user guide
\footnote{http://fsl.fmrib.ox.ac.uk/fsl/fslwiki/FIRST/UserGuide}, and
\freesurfer{} was run with the command \verb+recon-all -all+.

@@

As in \expadnixval, we find that similarity score increases with a greater
number of atlases or templates but quickly plateaus (Figure \ref{fig:FEP-xval}).
A maximum similarity score of \Sexpr{fep.mb.kappa.best} is found when using 9
atlases, 19 templates, ANTS registration, and majority vote label fusion.

@@

As \citet{Bland1986} noted, high correlation amongst measures of the same
quantity does not necessarily imply agreement (as correlation can be driven by a
large range in true values, for instance). Figure \ref{fig:FEP-Bland-Altman}
shows Bland-Altman plots illustrating the level of agreement of each method with
manual volumes. All methods show an obvious proportional bias: \freesurfer{}
and \fsl{} markedly underestimate smaller hippocampi and over-estimate large
hippocampi (the limits of agreement are between $-2482mm^3$ and $-784mm^3$, and
between $-1653mm^3$ and $79mm^3$, respectively), whereas both \mb{} methods show
a much less exaggerated, but conservative bias (limits of agreement between
$-67mm^3$ and $766mm^3$ when using FEP atlases, and between $-333mm^3$ and
$504mm^3$ when using Winterburn atlases). On average, \freesurfer{} and \fsl{}
overestimate hippocampal volume by about $1600 mm^3$ and $800 mm^3$,
respectively. In contrast, on average \mb{} underestimates volumes by about
$300mm^3$ when using FEP atlases and by about $80mm^3$ when using Winterburn
atlases (compared to the \psegs).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% FEP-XVAL: mean Kappa plot by atlases, templates, reg and voting method. 
<<fig:FEP-xval,fig.show='hide',dependson='load'>>=
  stderr <- function(x) sqrt(var(x)/length(x))
  ggplot(fep.xval.mean, aes(x=templates,y=k, colour=atlases)) + 
    stat_summary(fun.y=mean,geom='line', aes(y=k,weight=1,group=atlases)) + 
    stat_summary(fun.y=mean,geom='point',aes(y=k,weight=1,group=atlases)) + 
    stat_summary(fun.y=mean,
                 fun.ymin=function (x) mean(x) - stderr(x),
                 fun.ymax=function (x) mean(x) + stderr(x), 
                 geom='errorbar',aes(y=k, weight=1, group=atlases), 
                 linetype='solid', width=0.5) + 
    #stat_smooth(method='lm', formula=y~log(x)) + 
    scale_y_continuous(breaks=seq(0,1,0.01)) + 
    scale_colour_hue(name="Number of Atlases") +
    scale_linetype(name="Registration Method") +
    xlab( "Number of Templates" ) + 
    ylab( "Mean similarity (DSC)" ) + 
    theme(legend.direction = "horizontal", legend.position = "bottom") 
@

@@

<<fig:FEP-Bland-Altman,dependson='fig:FEP-vols',fig.width=10,fig.height=6,fig.show='hide'>>=
fep.volumes$diff = fep.volumes$manual - fep.volumes$volume
fep.volumes$mean = (fep.volumes$manual + fep.volumes$volume)/2
limits = ddply(fep.volumes, c("Method"), function (df) { 
  data.frame(
    y = mean(df$diff,na.rm=T) + c(-1.96,0,2) * sd(df$diff,na.rm=T) 
)})

ggplot(fep.volumes, aes(x= mean, y = diff)) + 
  facet_wrap( ~ Method, nrow=2, scales="fixed") + 
  geom_smooth(method="lm", formula=y~x, fullrange=T, se=F) + 
  scale_shape_discrete(solid=F,legend=F) + 
  geom_point(size=2,mapping=aes(shape=1)) +  
  geom_hline(aes(yintercept = y), linetype="dashed", colour="brown",
             data=limits, size=0.3, alpha=0.5) + 
  geom_text(aes(y = y, x=Inf, label=round(y,0)), colour="black", data = limits,
            hjust=1.1,vjust=-0.8, size=2.5) + 
  xlab(expression('Mean manual and computed volume ' (mm^3))) + 
  ylab(expression('manual - computed volume ' (mm^3))) + 
  scale_y_continuous(breaks=seq(-10000,4000, 1000)) + 
  scale_x_continuous(breaks=seq(0,10000, 1000)) + 
  theme(legend.direction = "horizontal", legend.position = "bottom", 
        axis.text.y = element_text(angle = 90, hjust=0.5), 
        axis.text = element_text(size = 8))
@

\begin{figure}
  \centering
  \begin{subfigure}[t]{0.48\textwidth} 
    \includegraphics[width=\textwidth]{figure/fig:FEP-xval}
    \caption{Dice's similarity score vs. atlas and template library size} 
    \label{fig:FEP-xval}
  \end{subfigure}%
  \hfill
  \begin{subfigure}[t]{0.48\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:FEP-vols} 
    \caption{Computed vs. manual hippocampus volume} 
    \label{fig:FEP-vols}
  \end{subfigure}
  \begin{subfigure}[b]{\textwidth}
    \includegraphics[width=\textwidth]{figure/fig:FEP-Bland-Altman}
    \caption{Bland-Altman plots of computed vs. manual hippocampus volume} 
    \label{fig:FEP-Bland-Altman}
  \end{subfigure}
  \caption{\textbf{First Episode Patient dataset validation.}
  All manual segmentation of the 81 subjects is done with the \prsnr-protocol.
  \mb{} uses \ants{} registration and majority vote label fusion. 
(\ref{fig:FEP-xval}) shows mean DSC score of \mb{} segmentations, as atlas and
  template library size is varied over a 5-fold validation. Error bars indicate
  standard error. 
(\ref{fig:FEP-vols}) shows segmentation volumes from \fsl,
  \freesurfer, \mb{} using the five Winterburn atlases (MAGeT-Winterburn), and
  \mb{} using five manually segmented FEP subjects as atlases (MAGeT-FEP). Linear
  fit lines are shown, with the shaded region showing standard error.
(\ref{fig:FEP-Bland-Altman}) shows the agreement between computed and
  manually volumes. The overall mean difference in volume, and limits of
  agreement ($\pm 1.96SD$) are shown by dashed horizontal lines. Linear fit lines
  are shown for each diagnosis group. Note, points below the mean difference
  indicate overestimation of the volume with respect to the manual volume, and
  vice versa.}
\end{figure}

@@

\paragraph{\adnidataset{} dataset}  %%%%%%%%%%%
The {\em \adnidataset{}} standardized dataset contains $1919$ images of subjects
diagnosed with Alzheimer's disease (AD), late-onset mild cognitive impairment
(LMCI), or who are identified as cognitively normal (CN). See Section
\ref{adnidataset} for study details, inclusion criteria and imaging
characteristics.

@@

\maper{} -- a \multiatlas segmentation tool , and \freesurfer{} hippocampal volumes for a subset of images were also
provided by ADNI, along with quality control data for each \freesurfer{}
segmentation (guidelines described in \citep{Hartig2010}).

\paragraph{Experiment details} %%%%%%%%%%
\mb{} was configured with an atlas library composed of the five Winterburn atlas
images (\expfepxval, section \ref{exp:fep-xval}) and segmentations. A template
library of 19 images were randomly selected from the target dataset of ADNI
subjects, and \ants{} registration and majority vote label fusion were used as
these were found to perform favourably in earlier experiments.

@@

%%%%%%%%%
% ADNI volumes by automated method
%%%%%%%%%
<<fig:ADNI-volumes-plot,dependson='load',fig.show='hide'>>=
melted=melt(means.complete, measure.vars=c("FS", "FSL", "MAPER","MAGeT"), 
            variable.name = "Method", value.name = "volume")
